Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-19 manuscript/manuscript.tex
Added Wild-Pfeiffer2008 reference in manuscript
Reviewed the text until Adaptive Discretization
Mostly simplified the wording, cut repeated text, and removed some
sections that are well explained in the references. There is enough new
material here that we don't need to go into details of the GLQ, which
has been extensively covered elsewhere.

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable 
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Geology and Geophysics, SOEST, University of Hawai‘i at
    Manoa, Honolulu, Hawaii, USA
    }

@@

Lithosphere's density variation with depth has been studied for almost a 
century.
On the first years \citet{Athy1930} obtained a decreasing exponential relation 
between both quantities by studying shale samples, while in the following 
decades other density functions have been proposed for different rock types 
\citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}. 
Furthermore, varying densities have been taken into account in forward and 
inversion gravity models, mostly applied to grabens and basins 
\citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, 
Welford2010}.

These forward gravity models have been developed for two or three dimensional 
bodies in cartesian coordinates that properly fit local scales applications. 
Nevertheless, the latest satellite missions have provided us gravity 
measurements with global coverage, which allow geologists and geophysics to 
perform modelling and interpretation on regional scales. This raises the 
importance of building forward models that reproduce the gravity anomalies for 
such scales.

The main issue that should be overcome is to take into account the curvature 
of the Earth, thus the forward model must be defined in spherical coordinates. 
A common way to achieve this is to discretize the Earth in spherical prisms 
known as tesseroids, which are defined by pairs of latitude, longitude and 
radial boundaries (see Figure \ref{fig:tesseroid-uieda}).
The calculation of the gravitational fields generated by an arbitrary 
tesseroid on any external point involves the resolution of volume 
integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion 
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre 
Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

In order to develop a forward model that computes the gravitational 
field generated by any tesseroid with arbitrary variable density, the 
Taylor series expansion is not well suited, because it would need the 
series expansion terms for each density function.
On the other hand, the GLQ consists in approximating the integral by a 
weighted sum of the effect of point masses located at scaled nodes of the 
Legendre polynomials, where the density variation is summarised into the 
weights.
Thus, this kind of numerical approximation allows to calculate the 
gravitational fields for any density function without any other 
information but the function itself.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A spherical prism, also known as  Tesseroid, in a geocentric spherical 
    coordinate system, with a computation point $P$ and its local north 
    oriented Cartesian coordinate system. Picture made by \citet{Uieda2015}.
}
\label{fig:tesseroid-uieda}
\end{figure}

\citet{Uieda2016} developed a forward gravitational model based on 
tesseroids with homogeneous densities using the GLQ approximation 
method.
The main issue of the GLQ is that the approximation becomes less 
accurate when the computation point is closer to the tesseroid or for 
smaller GLQ order \citep{Ku1977}.
Although the model developed by \citet{Uieda2016} uses a second order GLQ, it 
ensures the accuracy through a modified version of the adaptive 
discretization of \citet{Li2011}.
It consist in splitting the tesseroid into smaller ones when a certain 
distance-size ratio is lower than a predefined value ($D$) that controls the 
accuracy of the computation.
\citet{Uieda2016} have also objectively obtained standard values of $D$ 
for gravitational potential, gradients and tensor components comparing 
the numerical model with the fields generated by a spherical shell, 
which constitutes a special case of the volume integrals that has 
analytical solution \citep{LaFehr1991, Mikuska2006, Grombein2013}.
\todo[inline]{citar mas??}

On the other hand, a varying density inside an arbitrary tesseroid 
introduces a new kind of error into the numerical approximation.
This encouraged us to develop a new density-based discretization 
algorithm that divides the tesseroid into smaller ones taking into 
account the variation of the density function.
The number of subdivisions and thus the accuracy of the approximation 
will be controlled by predefined delta ratio $\delta$.

In summary, we present a new forward gravitational model that allows to 
compute the gravitational fields generated by any tesseroid with an 
arbitrary continuous density function on any external point.
It is based on the GLQ approximation whose accuracy is controlled by 
the modified distance-size discretization \citep{Uieda2016} 
and the new density-based discretization algorithms.

In order to ensure the accuracy of the numerical approximation we have 
determined predefined values for the $D$ and $\delta$ parameters by 
comparing the numerical approximation with the analytical solutions for 
a spherical shell with linear and exponential density, which had to be 
derived.

Furthermore, we developed a Python library that implements this forward 
model.
I makes use of some previously existing classes and functions 
from the Python library Fatiando a Terra for geophysical modelling, 
what saved us time and allowed us to focus on the new model.

Furthermore, we have the intention to include it into a future release 
of the Fatiando a Terra library in order to make its distribution and 
maintenance easier.
\todo[inline]{include this?}

In the following sections we will describe how the new algorithm works, its 
theoretical background, obtain default distance-size ratio and delta ratio 
values needed to get a numerical approximation with acceptable accuracy and 
finally show a simple application of the methodology by modelling the Nequen 
basin through a set of tesseroids with exponential density.

@@

\section{Theory}

The spherical prisms known as tesseroids are mass elements defined in a 
geocentric spherical coordinate system bounded by a pair of parallels, 
a pair of  meridians, and two concentric spherical surfaces.
We define an external computation point $P(r, \phi, \lambda)$ where the 
gravitational fields generated by the tesseroid are going to be calculated 
with respect to the local north oriented Cartesian coordinate system 
located at $P$.
In the special case of homogeneous density, the gravity potential, its 
gradient and the Marussi tensor components can be calculated through 
the integrals obtained by \citet{Grombein2013} \citep[for same notation 
as the one we will use, see][]{Uieda2016}.

In our case, we will assume that the tesseroid has an arbitrary 
variable density in depth, i.e. only depends on the radial spherical 
coordinate.
Thus, the integrals for the gravitational fields are slightly modified:

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesseroid-pot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}

@@

\noindent $\alpha, \beta \in \{x, y, z\}$, $\rho(r')$ is the density 
function that depends on the radial coordinate, $\delta_{\alpha\beta}$ 
is Kronecker's delta, $G = 6.674\times10^{-11}\, 
\text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and

@@

According to \citet[p.~390]{Hildebrand1987}, we can approximate each 
integral on equations \ref{eq:tesseroid-pot}, \ref{eq:tesseroid-grav} 
and \ref{eq:tesseroid-tensor} by applying a $N$th order GLQ, i.e. as 
a weighted sum of the integration kernel evaluated on the roots of the 
$N$th Legendre polynomial.
In the variable density tesseroid case the GLQ is applied in a similar 
way as \citet{Asgharzadeh2007} and \citet{Uieda2016} do, but in this 
scenario the density function must be also evaluated on the quadrature 
nodes:

@@

\noindent $(r_i, \phi_j, \lambda_k)$ are the GLQ nodes and $W_i^r$ are 
the quadrature weights \citep[see their definitions 
on][]{Hildebrand1987}.

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates 
the gravitational fields of a tesseroid as the ones generated by $N_\lambda 
N_\phi N_r$ point masses located at the nodes of the Legendre polynomials, in 
the same way as the GLQ works in the homogeneous density case, where $N_i$ ($i 
\in \{ \lambda, \phi, r \}$) are the orders of the quadrature for each 
integration.
Nevertheless, in this new variable density scenario, the information of 
the density function is summarised into the weights by taking into 
account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and 
show that this method, along with well suited discretization 
algorithms, approximates the gravitational fields with good accuracy.
\todo[inline]{no se si poner esta ultima oracion}

In a general way, the numerical error of the GLQ is controlled by the 
order of the quadrature: a higher GLQ order produces a more accurate 
approximation \citep{Hildebrand1987}.
This can also be understood in terms of the tesseroid problem: a higher 
quadrature order increases the number of point masses used to compute 
the gravitational fields and thus lowers the numerical error.

This reasoning could lead to a naive methodology in order to ensure 
accurate gravitational field computations: find the order of the 
quadrature that produces an acceptable numerical error.
Nevertheless, this simple proposal has a few problems: (a) it won't be 
efficient \citep{Li2011, Uieda2016} and (b) it couldn't be applied to 
the general case, i.e. the needed quadrature order probably won't be 
the same for every tesseroid, computation point or density function.

That's why we will keep the GLQ order fixed and make use of 
discretization algorithms.
Instead of uniformly increase the number of point masses that 
approximates a certain tesseroid, the discretization algorithms are 
intended to efficiently subdivide the tesseroid into smaller ones and 
apply the GLQ to each one of them.
These algorithms generally involve a discretization inequation that 
controls if the tesseroid should be subdivided or not, and a certain 
predefined parameter that works as a threshold for the discretization 
inequation.

On the variable density tesseroid problem we found two main error 
sources in the GLQ that should be addressed: (a) the distance between 
the tesseroid and the computation point, and (b) 
the variation of the density function.
The former one is also present in the homogeneous density tesseroid 
models, such as the one developed by \citet{Uieda2016}, and could be 
solved by applying a modified version of the adaptive discretization 
algorithm proposed by \citet{Li2011}.
Although for the latter we need to introduce a new density-based 
discretization algorithm that subdivides the tesseroid taking into 
account the variation of its density function.

\subsection{Modified Adaptive Discretization Algorithm}

>>>>

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Geology and Geophysics, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

The lithosphere's density variation with depth has been studied for close to a
century.
Over this time period,
several density-depth relations have been proposed for different rock types
\citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, depth-variable densities have been used in forward and
inverse gravity modeling, mostly applied to sedimentary basins
\citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001,
Welford2010}.
These forward gravity models have been developed for two or three dimensional
bodies in Cartesian coordinates, which limits the applications to local scales.
The advent of satellite gravimetry has provided gravity field
measurements with global coverage, enabling modelling and interpretation on regional and
global scales.
Hence, designing forward modeling methods that reproduce the gravity anomalies for
such scales is of high importance.

To take into account the curvature of the Earth, many global forward modeling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids (spherical prisms),
which are defined by pairs of latitude, longitude, and
radial boundaries (see Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by an arbitrary
tesseroid on any external point are given by volume
integrals that are generally approximated by numerically.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary density function can be included in the GLQ without
any change to the integration method.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the fields generated by a spherical shell.

We present a new algorithm for computing the gravitational fields generated by any
tesseroid with an arbitrary continuous density function on any external point.
It is based on the GLQ approximation and the adaptive discretization of
\citep{Uieda2016}, which we extend to include a new density-based discretization step.
In order to ensure the accuracy of the numerical approximation we have
determined optimal values for the controlling parameters by
comparing the numerical approximation with analytical solutions for
spherical shells.
Finally, we applied the methodology to model the Nequen basin, Argentina, using
tesseroids with an exponentially increasing density with depth.

@@

We define an external computation point $P(r, \phi, \lambda)$ in a geocentric spherical
coordinate system at a radius $r$, geocentric latitude $\phi$, and longitude $\lambda$
where the gravitational fields are going to be calculated.
The first and second derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first and second derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesseroid-pot}
\end{equation}

\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}

\noindent and

\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}

\noindent in which

@@

\noindent $\alpha, \beta \in \{x, y, z\}$, $\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant
and

@@

\subsection{Gauss-Legendre Quadrature integration}

Applying a $N$th order GLQ, we can approximate each integral in equations
\ref{eq:tesseroid-pot}, \ref{eq:tesseroid-grav} and \ref{eq:tesseroid-tensor} by a
weighted sum of the integration kernel evaluated on the roots of an $N$th order Legendre
polynomial \citep[p.~390]{Hildebrand1987}.
Unlike the homogeneous density case, the radial density function $\rho(r)$ must also be
included in the integration and evaluated on the Legendre polynomial roots (i.e.,
quadrature nodes).

@@

\noindent $(r_i, \phi_j, \lambda_k)$ are the GLQ nodes, and $W_i^r$ are
the quadrature weights.

\subsection{Adaptive Discretization}
